---
title: "A Replication of Karlan and List (2007)"
author: "Your Name"
date: today
callout-appearance: minimal # this hides the blue "i" icon on .callout-notes
---


## Introduction

Dean Karlan at Yale and John List at the University of Chicago conducted a field experiment to test the effectiveness of different fundraising letters. They sent out 50,000 fundraising letters to potential donors, randomly assigning each letter to one of three treatments: a standard letter, a matching grant letter, or a challenge grant letter. They published the results of this experiment in the _American Economic Review_ in 2007. The article and supporting data are available from the [AEA website](https://www.aeaweb.org/articles?id=10.1257/aer.97.5.1774) and from Innovations for Poverty Action as part of [Harvard's Dataverse](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/27853&version=4.2).

_to do: expand on the description of the experiment._

This project seeks to replicate their results.


## Data

### Description


```{r}
library(haven)

# Load the data from the current working directory
data <- read_dta("karlan_list_2007.dta")
list.files()
```
### Description
```{r}

#_todo: Read the data into R/Python and describe the data_
## Load libraries
library(haven)
library(dplyr)
library(ggplot2)

# Read the data
data <- read_dta("karlan_list_2007.dta")

# View structure of the dataset
str(data)

data %>%
  group_by(treatment) %>%
  summarise(
    response_rate = mean(gave),
    avg_donation = mean(amount),
    n = n()
  )

```

:::: {.callout-note collapse="true"}
### Variable Definitions

| Variable             | Description                                                         |
|----------------------|---------------------------------------------------------------------|
| `treatment`          | Treatment                                                           |
| `control`            | Control                                                             |
| `ratio`              | Match ratio                                                         |
| `ratio2`             | 2:1 match ratio                                                     |
| `ratio3`             | 3:1 match ratio                                                     |
| `size`               | Match threshold                                                     |
| `size25`             | \$25,000 match threshold                                            |
| `size50`             | \$50,000 match threshold                                            |
| `size100`            | \$100,000 match threshold                                           |
| `sizeno`             | Unstated match threshold                                            |
| `ask`                | Suggested donation amount                                           |
| `askd1`              | Suggested donation was highest previous contribution                |
| `askd2`              | Suggested donation was 1.25 x highest previous contribution         |
| `askd3`              | Suggested donation was 1.50 x highest previous contribution         |
| `ask1`               | Highest previous contribution (for suggestion)                      |
| `ask2`               | 1.25 x highest previous contribution (for suggestion)               |
| `ask3`               | 1.50 x highest previous contribution (for suggestion)               |
| `amount`             | Dollars given                                                       |
| `gave`               | Gave anything                                                       |
| `amountchange`       | Change in amount given                                              |
| `hpa`                | Highest previous contribution                                       |
| `ltmedmra`           | Small prior donor: last gift was less than median \$35              |
| `freq`               | Number of prior donations                                           |
| `years`              | Number of years since initial donation                              |
| `year5`              | At least 5 years since initial donation                             |
| `mrm2`               | Number of months since last donation                                |
| `dormant`            | Already donated in 2005                                             |
| `female`             | Female                                                              |
| `couple`             | Couple                                                              |
| `state50one`         | State tag: 1 for one observation of each of 50 states; 0 otherwise  |
| `nonlit`             | Nonlitigation                                                       |
| `cases`              | Court cases from state in 2004-5 in which organization was involved |
| `statecnt`           | Percent of sample from state                                        |
| `stateresponse`      | Proportion of sample from the state who gave                        |
| `stateresponset`     | Proportion of treated sample from the state who gave                |
| `stateresponsec`     | Proportion of control sample from the state who gave                |
| `stateresponsetminc` | stateresponset - stateresponsec                                     |
| `perbush`            | State vote share for Bush                                           |
| `close25`            | State vote share for Bush between 47.5% and 52.5%                   |
| `red0`               | Red state                                                           |
| `blue0`              | Blue state                                                          |
| `redcty`             | Red county                                                          |
| `bluecty`            | Blue county                                                         |
| `pwhite`             | Proportion white within zip code                                    |
| `pblack`             | Proportion black within zip code                                    |
| `page18_39`          | Proportion age 18-39 within zip code                                |
| `ave_hh_sz`          | Average household size within zip code                              |
| `median_hhincome`    | Median household income within zip code                             |
| `powner`             | Proportion house owner within zip code                              |
| `psch_atlstba`       | Proportion who finished college within zip code                     |
| `pop_propurban`      | Proportion of population urban within zip code                      |

::::


### Balance Test 

As an ad hoc test of the randomization mechanism, I provide a series of tests that compare aspects of the treatment and control groups to assess whether they are statistically significantly different from one another.


```{r}
# Load required libraries
library(haven)

# Load the dataset
data <- read_dta("karlan_list_2007.dta")

# Variables to test
vars <- c("mrm2", "freq", "years")

# Loop through each variable and perform t-test and regression
for (var in vars) {
  cat("\n==============================\n")
  cat("Variable:", var, "\n")

  # Remove missing values
  df_clean <- na.omit(data[, c(var, "treatment")])

  # T-test
  t_test <- t.test(df_clean[[var]] ~ df_clean$treatment, var.equal = FALSE)
  print(t_test)

  # Linear regression
  formula <- as.formula(paste(var, "~ treatment"))
  lm_model <- lm(formula, data = df_clean)
  print(summary(lm_model))
}

```

Table 1 shows that the treatment and control groups were balanced on key baseline characteristics. That’s important because:
1. It confirms that random assignment worked.
2. It rules out the possibility that observed treatment effects are due to pre-existing differences.


## Experimental Results

### Charitable Contribution Made

First, I analyze whether matched donations lead to an increased response rate of making a donation. 

```{r}
# Load necessary libraries
library(haven)
library(ggplot2)
library(dplyr)

# Read the data
data <- read_dta("karlan_list_2007.dta")

# Calculate donation proportions
donation_props <- data %>%
  group_by(treatment) %>%
  summarise(prop_donated = mean(gave, na.rm = TRUE)) %>%
  mutate(group = ifelse(treatment == 1, "Treatment", "Control"))

# Plot
ggplot(donation_props, aes(x = group, y = prop_donated, fill = group)) +
  geom_bar(stat = "identity", width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(prop_donated, accuracy = 0.1)), 
            vjust = -0.5, size = 5) +
  scale_fill_manual(values = c("Control" = "#87CEEB", "Treatment" = "#1E90FF")) +
  labs(
    title = "Proportion of People Who Donated",
    x = "Group",
    y = "Proportion Donated"
  ) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.05)) +  # Set limit if values are small
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

```

```{r}
# Load necessary libraries
library(haven)

# Load the data
data <- read_dta("karlan_list_2007.dta")

# Drop missing values
data_clean <- na.omit(data[, c("gave", "treatment")])

# T-test
t_test_result <- t.test(gave ~ treatment, data = data_clean, var.equal = FALSE)
print(t_test_result)

# Bivariate linear regression
lm_result <- lm(gave ~ treatment, data = data_clean)
summary(lm_result)

```

1. The coefficient on treatment is the marginal effect (in latent index units) of being in the treatment group.
2. A positive, significant coefficient supports the claim that the treatment increased the likelihood of donating.

```{r}
# Load necessary libraries
library(haven)

# Load the data
data <- read_dta("karlan_list_2007.dta")

# Drop missing values for the relevant variables
data_clean <- na.omit(data[, c("gave", "treatment")])

# Run the probit regression
probit_model <- glm(gave ~ treatment, family = binomial(link = "probit"), data = data_clean)

# View the summary
summary(probit_model)

```
1. The treatment effect is positive and statistically significant (p ≈ 0.0019).
2. This supports Table 3, Column 1 of the paper: being assigned to treatment increases the probability of donating — in latent terms.
3. The latent index increase of 0.087 suggests a meaningful shift in underlying propensity to donate.

### Differences between Match Rates

Next, I assess the effectiveness of different sizes of matched donations on the response rate.


```{r}
# Load necessary library
library(haven)

# Load the data
data <- read_dta("karlan_list_2007.dta")

# Filter to only treatment group and relevant variables
data_match <- subset(data, treatment == 1 & !is.na(ratio) & !is.na(gave))

# Create subsets by match ratio
gave_1to1 <- subset(data_match, ratio == 1)$gave
gave_2to1 <- subset(data_match, ratio == 2)$gave
gave_3to1 <- subset(data_match, ratio == 3)$gave

# T-tests
t_test_2vs1 <- t.test(gave_2to1, gave_1to1, var.equal = FALSE)
t_test_3vs1 <- t.test(gave_3to1, gave_1to1, var.equal = FALSE)

# Print results
t_test_2vs1
t_test_3vs1
```
This supports the authors comment 
1. The existence of a match offer boosts giving.
2. But increasing the generosity of the match (from 1:1 to 2:1 or 3:1) doesn’t significantly change donor behavior.


```{r}
# Load required library
library(haven)

# Load the dataset
data <- read_dta("karlan_list_2007.dta")

# Filter treatment group and non-missing values
data_match <- subset(data, treatment == 1 & !is.na(ratio) & !is.na(gave))

# Create dummy variables for match ratios
data_match$ratio1 <- ifelse(data_match$ratio == 1, 1, 0)
data_match$ratio2 <- ifelse(data_match$ratio == 2, 1, 0)
data_match$ratio3 <- ifelse(data_match$ratio == 3, 1, 0)

# Option 1: Regression using dummy variables
model_dummies <- lm(gave ~ ratio1 + ratio2 + ratio3, data = data_match)
summary(model_dummies)

# Option 2: Regression using ratio as a categorical variable
model_categorical <- lm(gave ~ factor(ratio), data = data_match)
summary(model_categorical)

```

1. The treatment group gave slightly more on average than the control group.
2. However, the p-value (0.055) is just above the conventional 0.05 threshold for statistical significance.
3. This means that we can't confidently say that treatment caused people to donate more — the observed difference could be due to chance.

```{r}
# Load the data
library(haven)
data <- read_dta("/Users/siddharthamysore/sid_site/blog/AB_Testing/karlan_list_2007.dta")

# Filter to treatment group with non-missing values
data_match <- subset(data, treatment == 1 & !is.na(ratio) & !is.na(gave))

# 1. Raw response rate differences
response_rates <- tapply(data_match$gave, data_match$ratio, mean)
diff_2vs1_raw <- response_rates["2"] - response_rates["1"]
diff_3vs2_raw <- response_rates["3"] - response_rates["2"]

# 2. Regression using factor(ratio)
model <- lm(gave ~ factor(ratio), data = data_match)
summary(model)

# Differences from fitted model (1:1 is reference group)
coef_2vs1 <- coef(model)["factor(ratio)2"]
coef_3vs1 <- coef(model)["factor(ratio)3"]
coef_3vs2 <- coef_3vs1 - coef_2vs1

# Print everything
print(response_rates)
cat("Raw Diff 2:1 - 1:1 =", diff_2vs1_raw, "\n")
cat("Raw Diff 3:1 - 2:1 =", diff_3vs2_raw, "\n")
cat("Fitted Diff 2:1 - 1:1 =", coef_2vs1, "\n")
cat("Fitted Diff 3:1 - 2:1 =", coef_3vs2, "\n")

```


### Size of Charitable Contribution

In this subsection, I analyze the effect of the size of matched donation on the size of the charitable contribution.

```{r}
# Load necessary library
library(haven)

# Load the data
data <- read_dta("/Users/siddharthamysore/sid_site/blog/AB_Testing/karlan_list_2007.dta")

# Drop missing values
data_clean <- na.omit(data[, c("amount", "treatment")])

# T-test
t_test_result <- t.test(amount ~ treatment, data = data_clean, var.equal = FALSE)
print(t_test_result)

# Bivariate linear regression
lm_result <- lm(amount ~ treatment, data = data_clean)
summary(lm_result)

```
1. The mean donation in the treatment group is slightly higher than in the control group (~$0.15 more).
2.But the p-value from both tests is just above 0.05, which means the difference is not statistically significant at the 95% confidence level.

```{r}
# Load necessary library
library(haven)

# Load the data
data <- read_dta("/Users/siddharthamysore/sid_site/blog/AB_Testing/karlan_list_2007.dta")

# Filter to donors only (amount > 0)
data_donors <- subset(data, amount > 0 & !is.na(treatment))

# T-test
t_test_donors <- t.test(amount ~ treatment, data = data_donors, var.equal = FALSE)
print(t_test_donors)

# Linear regression (donation amount ~ treatment)
lm_donors <- lm(amount ~ treatment, data = data_donors)
summary(lm_donors)

```
1. The t-test and linear regression both likely returned a non-significant difference in donation amounts between the treatment and control groups.
2. The regression coefficient for treatment was probably close to 0 or negative, and not statistically significant.


```{r}
# Load necessary library
library(haven)
library(ggplot2)
library(dplyr)

# Load the data
data <- read_dta("/Users/siddharthamysore/sid_site/blog/AB_Testing/karlan_list_2007.dta")

# Filter to only those who donated
data_donors <- subset(data, amount > 0 & !is.na(treatment))

# Calculate group means
group_means <- data_donors %>%
  group_by(treatment) %>%
  summarise(mean_amount = mean(amount))

# Plot for Treatment group
ggplot(data_donors[data_donors$treatment == 1, ], aes(x = amount)) +
  geom_histogram(binwidth = 2, fill = "skyblue", color = "black") +
  geom_vline(xintercept = group_means$mean_amount[group_means$treatment == 1], 
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(title = "Donation Amounts - Treatment Group", x = "Donation Amount", y = "Count") +
  theme_minimal()

# Plot for Control group
ggplot(data_donors[data_donors$treatment == 0, ], aes(x = amount)) +
  geom_histogram(binwidth = 2, fill = "lightgreen", color = "black") +
  geom_vline(xintercept = group_means$mean_amount[group_means$treatment == 0], 
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(title = "Donation Amounts - Control Group", x = "Donation Amount", y = "Count") +
  theme_minimal()

```

## Simulation Experiment

As a reminder of how the t-statistic "works," in this section I use simulation to demonstrate the Law of Large Numbers and the Central Limit Theorem.

Suppose the true distribution of respondents who do not get a charitable donation match is Bernoulli with probability p=0.018 that a donation is made. 

Further suppose that the true distribution of respondents who do get a charitable donation match of any size  is Bernoulli with probability p=0.022 that a donation is made.

### Law of Large Numbers

```{r}
# Load necessary library
library(haven)

# Load the data
data <- read_dta("/Users/siddharthamysore/sid_site/blog/AB_Testing/karlan_list_2007.dta")

# Filter to only positive donations
donations_control <- subset(data, treatment == 0 & amount > 0)$amount
donations_treatment <- subset(data, treatment == 1 & amount > 0)$amount

# Set seed for reproducibility
set.seed(42)

# Simulate: 100,000 draws from control, 10,000 from treatment
sim_control <- sample(donations_control, size = 100000, replace = TRUE)
sim_treatment <- sample(donations_treatment, size = 10000, replace = TRUE)

# Sample 10,000 control values to match treatment sample
sim_control_sample <- sample(sim_control, size = 10000, replace = FALSE)

# Calculate differences
differences <- sim_treatment - sim_control_sample

# Cumulative average
cumulative_avg_diff <- cumsum(differences) / seq_along(differences)

# True difference in means
true_diff <- mean(donations_treatment) - mean(donations_control)

# Plot
plot(cumulative_avg_diff, type = "l", col = "blue", lwd = 2,
     main = "Cumulative Average of Simulated Differences",
     xlab = "Simulation Index", ylab = "Cumulative Average Difference")
abline(h = true_diff, col = "red", lty = 2)
legend("bottomright", legend = c("Cumulative Average", "True Difference"),
       col = c("blue", "red"), lty = c(1, 2), lwd = 2)

```

Yes — the cumulative average approaches the true difference in means.
This illustrates the Law of Large Numbers: with more data, your estimate becomes more stable and closer to the true population value.
It also reinforces the power of randomization and repeated sampling — even if individual draws are noisy, the average across many replicates becomes reliable.

### Central Limit Theorem

```{r}
# Load necessary library
library(haven)

# Load the dataset
data <- read_dta("/Users/siddharthamysore/sid_site/blog/AB_Testing/karlan_list_2007.dta")

# Filter to positive donation amounts
donations_control <- subset(data, treatment == 0 & amount > 0)$amount
donations_treatment <- subset(data, treatment == 1 & amount > 0)$amount

# Set seed for reproducibility
set.seed(42)

# Function to simulate 1000 average differences at a given sample size
simulate_diffs <- function(n, reps = 1000) {
  replicate(reps, {
    sample_control <- sample(donations_control, size = n, replace = TRUE)
    sample_treat <- sample(donations_treatment, size = n, replace = TRUE)
    mean(sample_treat) - mean(sample_control)
  })
}

# Sample sizes to test
sample_sizes <- c(50, 200, 500, 1000)
diffs_list <- lapply(sample_sizes, simulate_diffs)

# Plot the histograms
par(mfrow = c(2, 2))  # 2x2 layout
for (i in 1:4) {
  hist(diffs_list[[i]], main = paste("Sample Size =", sample_sizes[i]),
       xlab = "Avg Difference (Treatment - Control)", col = "skyblue", border = "black")
  abline(v = 0, col = "red", lwd = 2, lty = 2)
}
```
At small sample sizes (e.g., 50):

The distribution of average differences is wide and noisy.
The red line at zero often appears in the middle — suggesting a high chance of not detecting a difference.
As sample size increases (200 → 500 → 1000):

The distribution becomes narrower.
It starts to center away from zero, and zero moves to the tail of the distribution.
This reflects increasing statistical power — you’re more likely to detect the true effect as the sample grows.

```